Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.

### 2024-01-25 manuscript/main.tex
finalize figs for draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.},
  pdfauthor={Alex Barth*; Leo Blanco-Berical; Rod Johnson; Joshua Stone},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

Rhizaria are an extremely diverse super-group of single celled organisms
consisting of several phyla including retaria (foraminifera and
radiolaria) and cercozoa. These organisms exist in a wide range of
habitats and are widely represented in plankton communities throughout
the global ocean. While the taxonomy of these organisms has recently
undergone several reclassifications (Wiley, 2022), their presence in
ocean ecosystems has been long known to oceanographers. Some of the
earliest records of their existence are from oceanographic expeditions
in the 19th century (Haekel, 1887). Rhizaria are unique members of the
plankton and protist community because they can reach large sizes (up to
several mm in diameter) and they construct intricate mineral skeletons
out of either silica, strontium, or calcium carbonate (Kimoto, 2015;
Nakamura and Suzuki, 2015; Suzuki and Not, 2015; Wiley, 2022). Despite
their noticeable morphology and global distribution, Rhizaria were
largely understudied throughout the 20th century. The bulk of modern
plankton research has focused hard-bodied crustacea which are
numerically dominant and easily sampled with nets and preservatives.
Fragile organisms like Rhizaria, were difficult to adequately study as
they can be destroyed through standard zooplankton sampling techniques.
A number of studies in the late 1900s did employ alternative techniques
to quantify Rhizaria including diaphragm pumps (Michaels, 1988) or
blue-water SCUBA collections (Bijma et al., 1990; Caron et al., 1995;
Caron and Be, 1984). However, the bulk of Rhizaria research was
constrained to sediment traps or paleontological studies of sediment
(Boltovskoy et al., 1993; Takahashi et al., 1983). Only recently has the
advent of molecular techniques and in-situ imaging tools ignited a
renewed focus on Rhizaria in pelagic ecosystems (Caron, 2016).

The wave of new data on Rhizaria has facilitated an improved
understanding of the significance in ocean ecosystem functions. Firstly,
taxonomists have been able to greatly refine the understanding of
evolutionary relationships amongst these diverse protists (Aurahs et
al., 2009; Biard et al., 2015; Cavalier-Smith et al., 2018; Decelle et
al., 2013, 2012; rev by Wiley, 2022). DNA metabarcoding studies have
revealed insights into the distributional patterns Llopis Monferrer et
al. (2022), ecological relationships (Decelle et al., 2012; Nakamura et
al., 2023), and contribution to biogeochemical fluxes (Guidi et al.,
2016; Gutierrez-Rodriguez et al., 2019). Transcriptomic and proteomic
approaches also have been used to quantify rhizarian contribution to
community metabolism (Cohen et al., 2023). Yet, despite the excellent
taxonomic resolution provided by molecular approaches, they do not
provide a truly quantitative metric for estimating Rhizarian abundance
or biomass. In-situ imaging tools however, offer the ability to observe
organisms in the natural state and quantify their abundance (Barth and
Stone, in review; Ohman, 2019). While there were early applicaitons of
imaging tools to document Rhizaria (Dennett et al., 2002), Biard et al.
(2016)'s report from a global imaging dataset highlighted the importance
of Rhizaria to the total standing stock of marine carbon. Due to their
large sizes, ability to concentrate smaller particles (ballasting), and
the unique structure of their mineral skeletons, Rhizaria have the
potential to massively influence ocean biogeochemical cycling. A number
of studies have made large advances in estimating the contribution of
Rhizaria to ocean cycling of carbon (Gutierrez-Rodriguez et al., 2019;
Ikenoue et al., 2019; Lampitt et al., 2009; Stukel et al., 2018), silica
(Biard et al., 2018; Llopis Monferrer et al., 2021), and strontium
(Decelle et al., 2013). Still, Rhizarian ecological roles are not well
understood (Wiley, 2022). This is a major challenge as it is critical to
understand the ecological role of plankton to fully incorporate them
into biological oceanographic models.

The ecological role of Rhizaria in plankton communities is complicated
due to the fact different taxa can exhibit every different trophic
modes. As zooplankton, rhizaria are predominately heterotrophic (Wiley,
2022), yet their feeding modes can be quite varied. Phaeodarians (family
Cercozoa) are largely thought to be flux-feeders, collecting and feeding
on sinking particles (Nakamura and Suzuki, 2015; Stukel et al., 2019).
Alternatively, Retaria can be either exclusively heterotrophic or
mixotrophic, utilizing photosynthetic algal symbionts (Anderson, 2014;
Decelle et al., 2015). Mixotrophic foraminifera host a variety of
endosymbiont partners (Decelle et al., 2015; Lee, 2006), which are
thought to support early and adult life stages and contribute to total
primary productivity (Kimoto, 2015). Still, foraminifera are omnivorous,
possibly even predominately carnivorous with several studies suggesting
that they can be effective predators (Anderson and Bé, 1976; Gaskell et
al., 2019), majoritively consuming live copepods (Caron and Be, 1984).
Radiolaria have several lineages all which have some taxa who are well
known to host symbionts (Biard, 2022). Amongst radiolarians, arguably
the most widespread are Collodaria who can be either large solitary
cells or form massive colonies, up to several meters in length (Swanberg
and Anderson, 1981). All known Collodaria species host dinoflagellate
symbionts (Biard, 2022) and can contribute substantially to primary
productivity, particularly in oligotrophic ocean regions (Caron et al.,
1995; Dennett et al., 2002). This Collodaria-symbiont association has
been suggested as a reason for their high abundances throughout the
photic zone of oligotrophic environments globally (Biard et al., 2017,
2016). A few Acantharean (Radiolarian order) clades host algal symbionts
(Biard, 2022; Decelle et al., 2012), notably with two clades forming an
exclusive relationship with Phaeocystis. However, globally Acantharea
are less abundant than Collodaria (Wiley, 2022) and contribute less to
total primary productivity (Michaels et al., 1995). This may be due to
the fact several clades of Acantharea are cyst-forming and strictly
heterotrophs (Biard, 2022; Decelle et al., 2013). Furthermore, Mars
Brisbin et al. (2020) documented apparent predation behavior in
Acantharea near the surface, suggesting that there may be a large
reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among
Rhizarian taxa, it is reasonable to expect some form of niche
partitioning. A number of studies do suggest evidence for vertical
zonation between Rhizaria groups according to various trophic
strategies. Taxa-specific studies of radiolarians suggest they may be
restricted to the euphotic zone (Boltovskoy, 2017; Michaels, 1988).
Although some studies report Acantharea in deeper waters (Decelle et
al., 2013; Gutiérrez-Rodríguez et al., 2022). Phaeodarians
alternatively, are generally found in the mesopelagic where
photosynthesis cannot occur but they can feed on sinking particles
(Stukel et al., 2018). In an imaging-based study of the whole Rhizaria
community, Biard and Ohman (2020) noted clear patterns in vertical
zonation which largely corresponded to different trophic roles. In the
oligotrophic ocean, Blanco-Bercial et al. (2022) also noted that the
protist community, including Rhizaria partition along an autotroph and
mixotroph to heterotroph gradient with increasing depth in the water
column. Yet, few studies have made direct attempts to relate rhizaria
abundances to abiotic environmental factors (Biard and Ohman, 2020). In
part, this is due to the fact few studies have been able to sample
Rhizaria in the same location over a consistent timeframe (Boltovskoy et
al., 1993; Gutiérrez-Rodríguez et al., 2022; Hull et al., 2011; Michaels
et al., 1995; Michaels, 1988). Furthermore, no studies have utilized
imaging, arguably the best method for quantifying rhizaria, consistently
throughout the full mesopelagic. Given this lack of information, there
are many unknowns with respect to Rhizarian ecology, seasonality and
phenology across different groups.

In this study, we present a comprehensive assessment of large Rhizaria
measured for multiple months (greater than 1 year) using an in-situ
imaging approach. With this dataset, we address two critical aims. 1)
Quantification of large Rhizaria throughout the epipelagic (0-200m) and
mesopelagic (200-1000m) over the course of an annual cycle. These data
were collected in the Sargasso Sea, and represents the first study of
its kind in an oligotrophic system. 2) We aim to test the hypothesis
that Rhizaria exhibit niche partitioning according to trophic roles.
This hypothesis makes several predictions, including vertical zonation,
as seen in prior studies, but also that abiotic variables related to
trophic strategy will explain abundance patterns. Specifically,
autotrophic/mixotrophic taxa will correspond to variables related to
autotrophy (chl-a concentration, primary productivity, high \(O_2\) )
and other rhizaria will correspond to factors which promote heterotrophy
(particle concentration, flux, and low \(O_2\)).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

@@

\usepackage{hyphenat} % don't hyphenat titles
\usepackage{graphicx}
\usepackage{lineno}\linenumbers

% Add any tex header commands here
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.},
  pdfauthor={Alex Barth*; Leo Blanco-Berical; Rod Johnson; Joshua Stone},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
    \begin{titlepage}
% This is a combination of Pandoc templating and LaTeX
% Pandoc templating https://pandoc.org/MANUAL.html#templates
% See the README for help

\raggedleft % left align the title page
\rule{1pt}{\textheight} % Vertical line
\hspace{0.05\textwidth} % Whitespace between the vertical line and title page text
% Adjust num before \textwidth to move the block left or right
\begin{minipage}[b][\textheight][s]{0.85\textwidth}

\raggedright
% Title and subtitle
{\large\bfseries\nohyphens{Rhizaria in the oligotrophic exhibit clear
temporal and vertical variability.}}\\[2\baselineskip]

  % Authors	
% This hairy bit of code is just to get "and" between the last 2
% authors. See below if you don't need that
 {\large{Alex Barth*}}{\textsuperscript{1}}%
%
, 
 {\large{Leo Blanco-Berical}}{\textsuperscript{2}}%
%
, 
 {\large{Rod Johnson}}{\textsuperscript{3}}%
%
%
{ and \large{Joshua Stone}}%
{\textsuperscript{1}}%
%

% This is how to do it if you don't need the "and"

  %%%%%% Affiliations
\vspace{2\baselineskip}

\hangindent=1em
\hangafter=1
%
{1}.~{University of South Carolina}%
%
, %
{Biological Sciences}%
%
%
, %
{700 Sumter St 401, Columbia, SC 29208}%
%
\par\hangindent=1em\hangafter=1%
%
{2}.~{Arizone State University}%
%
%
\par\hangindent=1em\hangafter=1%
%
{3}.~{Bermuda Institute of Ocean Sciences}%
%
%
, %
{Bermuda GE 01, UK}%
%

  %%%%%% Correspondence
\vspace{1\baselineskip}

  %use \vfill instead to get the space to fill flexibly	
%\vspace{0.25\textheight} % Whitespace between the title block and the publisher

\vfill

%%%%%% Cover image

  % Whitespace between the title block and the tagline
\vspace{0.1\textheight}

\end{minipage}  \end{titlepage}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

Rhizaria are an extremely diverse super-group of single celled organisms
consisting of several phyla including retaria (foraminifera and
radiolaria) and cercozoa. These organisms exist in a wide range of
habitats and are widely represented in plankton communities throughout
the global ocean. While the taxonomy of these organisms has recently
undergone several reclassifications (Biard, 2022a), their presence in
ocean ecosystems has been long known to oceanographers. Some of the
earliest records of their existence are from oceanographic expeditions
in the 19th century (Haekel, 1887). Rhizaria are unique members of the
plankton and protist community because they can reach large sizes (up to
several mm in diameter) and they construct intricate mineral skeletons
out of either silica, strontium, or calcium carbonate (Biard, 2022a;
Kimoto, 2015; Nakamura and Suzuki, 2015; Suzuki and Not, 2015). Despite
their noticeable morphology and global distribution, Rhizaria were
largely understudied throughout the 20th century. The bulk of modern
plankton research has focused hard-bodied crustacea which are
numerically dominant and easily sampled with nets and preservatives.
Fragile organisms like Rhizaria, were difficult to adequately study as
they can be destroyed through standard zooplankton sampling techniques.
A number of studies in the late 1900s did employ alternative techniques
to quantify Rhizaria including diaphragm pumps (Michaels, 1988) or
blue-water SCUBA collections (Bijma et al., 1990; Caron et al., 1995;
Caron and Be, 1984). However, the bulk of Rhizaria research was
constrained to sediment traps or paleontological studies of sediment
(Boltovskoy et al., 1993; Takahashi et al., 1983). Only recently has the
advent of molecular techniques and in-situ imaging tools ignited a
renewed focus on Rhizaria in pelagic ecosystems (Caron, 2016).

The wave of new data on Rhizaria has facilitated an improved
understanding of the significance in ocean ecosystem functions. Firstly,
taxonomists have been able to greatly refine the understanding of
evolutionary relationships amongst these diverse protists (Aurahs et
al., 2009; Biard et al., 2015; Cavalier-Smith et al., 2018; Decelle et
al., 2013, 2012; rev by Biard, 2022a). DNA metabarcoding studies have
revealed insights into the distributional patterns Llopis Monferrer et
al. (2022), ecological relationships (Decelle et al., 2012; Nakamura et
al., 2023), and contribution to biogeochemical fluxes (Guidi et al.,
2016; Gutierrez-Rodriguez et al., 2019). Transcriptomic and proteomic
approaches also have been used to quantify rhizarian contribution to
community metabolism (Cohen et al., 2023). Yet, despite the excellent
taxonomic resolution provided by molecular approaches, they do not
provide a truly quantitative metric for estimating Rhizarian abundance
or biomass. In-situ imaging tools however, offer the ability to observe
organisms in the natural state and quantify their abundance (Barth and
Stone, in review; Ohman, 2019). While there were early applicaitons of
imaging tools to document Rhizaria (Dennett et al., 2002), Biard et al.
(2016)'s report from a global imaging dataset highlighted the importance
of Rhizaria to the total standing stock of marine carbon. Due to their
large sizes, ability to concentrate smaller particles (ballasting), and
the unique structure of their mineral skeletons, Rhizaria have the
potential to massively influence ocean biogeochemical cycling. A number
of studies have made large advances in estimating the contribution of
Rhizaria to ocean cycling of carbon (Gutierrez-Rodriguez et al., 2019;
Ikenoue et al., 2019; Lampitt et al., 2009; Stukel et al., 2018), silica
(Biard et al., 2018; Llopis Monferrer et al., 2021), and strontium
(Decelle et al., 2013). Still, Rhizarian ecological roles are not well
understood (Biard, 2022a). This is a major challenge as it is critical
to understand the ecological role of plankton to fully incorporate them
into biological oceanographic models.

The ecological role of Rhizaria in plankton communities is complicated
due to the fact different taxa can exhibit every different trophic
modes. As zooplankton, rhizaria are predominately heterotrophic (Biard,
2022a), yet their feeding modes can be quite varied. Phaeodarians
(family Cercozoa) are largely thought to be flux-feeders, collecting and
feeding on sinking particles (Nakamura and Suzuki, 2015; Stukel et al.,
2019). Alternatively, Retaria can be either exclusively heterotrophic or
mixotrophic, utilizing photosynthetic algal symbionts (Anderson, 2014;
Decelle et al., 2015). Mixotrophic foraminifera host a variety of
endosymbiont partners (Decelle et al., 2015; Lee, 2006), which are
thought to support early and adult life stages and contribute to total
primary productivity (Kimoto, 2015). Still, foraminifera are omnivorous,
possibly even predominately carnivorous with several studies suggesting
that they can be effective predators (Anderson and Bé, 1976; Gaskell et
al., 2019), majoritively consuming live copepods (Caron and Be, 1984).
Radiolaria have several lineages all which have some taxa who are well
known to host symbionts (Biard, 2022b). Amongst radiolarians, arguably
the most widespread are Collodaria who can be either large solitary
cells or form massive colonies, up to several meters in length (Swanberg
and Anderson, 1981). All known Collodaria species host dinoflagellate
symbionts (Biard, 2022b) and can contribute substantially to primary
productivity, particularly in oligotrophic ocean regions (Caron et al.,
1995; Dennett et al., 2002). This Collodaria-symbiont association has
been suggested as a reason for their high abundances throughout the
photic zone of oligotrophic environments globally (Biard et al., 2017,
2016). A few Acantharean (Radiolarian order) clades host algal symbionts
(Biard, 2022b; Decelle et al., 2012), notably with two clades forming an
exclusive relationship with Phaeocystis. However, globally Acantharea
are less abundant than Collodaria (Biard, 2022a) and contribute less to
total primary productivity (Michaels et al., 1995). This may be due to
the fact several clades of Acantharea are cyst-forming and strictly
heterotrophs (Biard, 2022b; Decelle et al., 2013). Furthermore, Mars
Brisbin et al. (2020) documented apparent predation behavior in
Acantharea near the surface, suggesting that there may be a large
reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among
Rhizarian taxa, it is reasonable to expect some form of niche
partitioning. A number of studies do suggest evidence for vertical
zonation between Rhizaria groups according to various trophic
strategies. Taxa-specific studies of radiolarians suggest they may be
restricted to the euphotic zone (Boltovskoy, 2017; Michaels, 1988).
Although some studies report Acantharea in deeper waters (Decelle et
al., 2013; Gutiérrez-Rodríguez et al., 2022). Phaeodarians
alternatively, are generally found in the mesopelagic where
photosynthesis cannot occur but they can feed on sinking particles
(Stukel et al., 2018). In an imaging-based study of the whole Rhizaria
community, Biard and Ohman (2020) noted clear patterns in vertical
zonation which largely corresponded to different trophic roles. In the
oligotrophic ocean, Blanco-Bercial et al. (2022) also noted that the
protist community, including Rhizaria partition along an autotroph and
mixotroph to heterotroph gradient with increasing depth in the water
column. Yet, few studies have made direct attempts to relate rhizaria
abundances to environmental factors (Biard and Ohman, 2020). In part,
this is due to the fact few studies have been able to sample Rhizaria in
the same location over a consistent timeframe (Boltovskoy et al., 1993;
Gutiérrez-Rodríguez et al., 2022; Hull et al., 2011; Michaels et al.,
1995; Michaels, 1988). Furthermore, no studies have utilized imaging,
arguably the best method for quantifying rhizaria, consistently
throughout the full mesopelagic. Given this lack of information, there
are many unknowns with respect to Rhizarian ecology, seasonality and
phenology across different groups.

In this study, we present a comprehensive assessment of large Rhizaria
measured for multiple months (greater than 1 year) using an in-situ
imaging approach. With this dataset, we address two critical aims. 1)
Quantification of large Rhizaria throughout the epipelagic (0-200m) and
mesopelagic (200-1000m) over the course of an annual cycle. These data
were collected in the Sargasso Sea, and represents the first study of
its kind in an oligotrophic system. 2) We aim to test the hypothesis
that Rhizaria exhibit niche partitioning according to trophic roles.
This hypothesis makes several predictions, including vertical zonation,
as seen in prior studies, but also that environmental variables related
to trophic strategy will explain abundance patterns. Specifically,
autotrophic/mixotrophic taxa will correspond to variables related to
autotrophy (chl-a concentration, primary productivity, high \(O_2\) )
and other rhizaria will correspond to factors which promote heterotrophy
(particle concentration, flux, and low \(O_2\)).

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{oceanographic-sampling}{%
\subsection{Oceanographic Sampling}\label{oceanographic-sampling}}

Data were collected in collaboration with the Bermuda Atlantic
Time-Series Study (Lomas et al., 2013; Michaels and Knap, 1996) on board
the R/V Atlantic Explorer. Cruises were conducted at approximately
monthly intervals. Rhizaria individuals were sampled using the
Underwater Vision Profiler 5 {[}UVP5; Picheral et al. (2010){]}, a tool
which is well established to accurately quantify large Rhizaria (Barth
and Stone, 2022; Biard et al., 2016; Biard and Ohman, 2020; Drago et
al., 2022; Llopis Monferrer et al., 2022; Panaïotis et al., 2023; Stukel
et al., 2019; Stukel et al., 2018). The UVP5 was mounted to the sampling
rosette and collected data autonomously on routine casts, from which
only the downcast data are utilized. The UVP5 was deployed from
June-September 2019 then from October 2020 - January 2022. Casts were
filtered to only include data collected in the BATS region, far offshore
of Bermuda into the Sargasso Sea (Supplemental Figure 1). In general,
casts extended to either 200m, 500m, or 1200m deep, with a few extended
into the bathypelagic (4500m). However, Rhizaria were only typically
found in large abundances throughout the epipelagic and mesopelagic
zones. As such, we limit this study to results from the upper 1000m of
the water column.

A variety of biotic and abiotic data were collected during each BATS
cruise. Briefly, we will explain the data utilized in this study. The
UVP5 provided particle count data at a high-frequency from each cast.
Particle concentration was calculated from this data for all particles
184\(\mu m\) - 450\(\mu m\). The lower size range was set by what could
be reliably sampled by the UVP5's pixel resolution (\textgreater2px) and
the upper size range is representative of a potential prey field for
mesozooplankton (Whitmore and Ohman, 2021). Salinity, temperature,
Dissolved Oxygen (DO), and in-situ chlorophyll fluorescence were
measured at high-frequencies on each cast, attached to the same CTD
rosette as the UVP. On certain casts, niskin bottles were used to
collect bacterial abundance estimates (via epifluorescence microscopy)
as well as measure inorganic nutrients (\(NO_3\), and \(Si\) as
silicate/sicilic acid) at discrete depths. On each cruise, flux
estimates of total mass, carbon, and nitrogen were also collected using
sediment traps (in the present study we utilize flux to the mesopelagic
as the flux at 200m). Also primary productivity was estimated through
measuring \(C^{14}\) uptake rates from in-situ incubations. Full
descriptions of the BATS sampling program, methods, and data are
available freely online
\href{https://bats.bios.asu.edu/bats-data/}{(https://bats.bios.asu.edu/bats-data/)}
and reviewed in (Lomas et al., 2013).

Environmental data were processed in a variety of ways to match the
format of the Rhizaria abundance estimates (See below). CTD data were
collected at higher frequency than the UVP, so these data were averaged
within matching bins to the UVP5 data. Data from niskin bottles were
first linearly interpolated with a 1m resolution for each cast where
data were available. Then, because data were collected at different
frequencies over the course of a cruise, the interpolated data were
averaged across casts, then averaged into matching UVP5-sized bins.
Primary productivity estimates were totaled within the euphotic zone to
represent a ``total euphotic productivity''. Finally the data from the
sediment trap deployed to 200m was used to represent ``flux to the
mesopelagic''.

\hypertarget{rhizaria-imaging-processing-and-quantification}{%
\subsection{Rhizaria imaging processing and
quantification}\label{rhizaria-imaging-processing-and-quantification}}

Individual vignettes of Rhizaria images were identified using the
classification platform Ecotaxa {[}cite{]}. Data were pre-sorted
according to a . Taxonomic classification were done based on morphology
exclusively. While there are sparse taxonomic guides for in-situ images
of rhizaria, identification largely relied on descriptions in (Nakamura
and Suzuki, 2015; Suzuki and Not, 2015; and Biard and Ohman, 2020).
Using the aforementioned sources and publicly available ecotaxa
projects, I constructed a guide which can be found here:
\url{https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf}.
Broadly, Rhizaria were classified as Foraminifera, Radiolarians
(Acantharea or Collodaria), or a variety of Phaeodarian families (Figure
1). When identification could not be confidently made between a few
candidate taxa, a less specific label was used. As a result, we have
data from ``unidentified Rhizaria'', which typically were vignettes not
distinguishable between Aulacanthidae or Acantharea or ``unidentified
Phaeodaria'', which are clearly phaeodaria but not distinguishable into
a family.

\begin{figure}

{\centering \includegraphics{images/01_taxa.pdf}

}

\caption{Example images of different Rhizaria taxa. 4mm scale bar shown
in lower right. All vignettes are same scale}

\end{figure}

The UVP5 samples at \textasciitilde15Hz rate as it descends the water
column and records the exact position of each particle larger than
600\(\mu m\). However, identified rhizaria ranged from a 934\(\mu m\)
Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To
confirm that the UVP5 was sampling adequately across all size ranges, an
NBSS slope was constructed to identify a drop-off which would indicate
poor-sampling at the small size range (Barth and Stone, in review;
Lombard et al., 2019) . However, it was evident from this analysis that
all size ranges were adequately sampled across the size range
(Supplemental Figure 2) so no data were excluded. The UVP5 reports the
exact depth at which a particle is recorded, however to estimate
abundance, observations must be binned over fixed depth intervals. In
our deployments had variable descent depths and speeds with more casts
descending to 500m than 1000m and descents quicker through the
epipelagic than the mesopelagic {[}see Barth and Stone (2022) for an
extended discussion of UVP5 data processing). For the present study,
Rhizarian abundances were estimated in 25m vertical bins, which offer a
moderate sampling volume (average 0.948\(m^3\) in the epipelagic and
0.589\(m^3\) in the mesopelagic) while still maintaining ecologically
relevant widths. However, concentrations in a 25m bin would need to be
greater than 2.428 indv. \(m^{-3}\) and 3.912 indv. \(m^{-3}\), in the
epipelagic and mesopelagic respectively, to fall below a 10\%
non-detection risk (Barth and Stone, in review; Benfield et al., 1996).
Because we typically observed many rhizaria taxa below these
concentrations, we present the 25m binned data to visualize broad-scale
average distributions. For quantifying and modelling Rhizaria
abundances, we present integrated abundance estimates, with each cast.
Due to the variable descent depths of the UVP, data are categorized as
epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic
(500-1000m). The average sampling volume integrated through these
regions were 7.59\(m^2\), 7.06\(m^2\), and 11.77\(m^2\), with
non-detection thresholds at 0.30 indv. \(m^{-2}\), 0.33 indv.
\(m^{-2}\), and 0.20 indv. \(m^{-2}\) respectively. All UVP data
processing was done using the \texttt{EcotaxaTools} package in R (Barth
2023).

\hypertarget{modelling-environmental-controls-of-rhizarian-abundance}{%
\subsection{Modelling environmental controls of Rhizarian
Abundance}\label{modelling-environmental-controls-of-rhizarian-abundance}}

Generalized Additive Models (GAMs) were used to assess the relationship
between integrated Rhizarian abundance and different environmental
factors. GAMs offer the ability to model non-linear and non-monotonic
relationships, which can be particularly useful in assessing ecological
relationships (Wood, 2017) and have been successfully applied to
Rhizarian ecology (Biard and Ohman, 2020). The \texttt{mgcv} package
(Wood, 2001) was used to construct models relating environmental
parameters to each taxonomic group's integrated abundance estimates from
each cast. To select the most parsimonious model for each analysis, a
backwards step-wise approach was taken. First, a full model was fit
using any term which may be ecologically relevant. Terms were fit using
maximum likelihood with a double penalty approach on unnecessary smooths
(Marra and Wood, 2011). The smoothness parameter was restricted (k = 6)
to prevent overfitting the models. At each iteration of the backwards
step-wise procedure, the model term with the lowest F score (least
statistically significant) was removed. This was repeated until all
model terms were statistically significant or the \(R^2_{adj}\) was
substantially reduced. Models were fit for each region; epipelagic,
upper mesopelagic, and lower mesopelagic. In cases where observations
were too sparse for a given taxonomic grouping, models were not run.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{environmental-variability}{%
\subsubsection{Environmental
Variability}\label{environmental-variability}}

The BATS sampling region is southeast of Bermuda, situated in the North
Sargasso Sea and the North Atlantic oligotrophic ocean gyre. Due to the
sampling location, while the environmental conditions are generally low
in variation and oligotrophic, there is some seasonality and
considerable influence from mesoscale eddies (Lomas et al., 2013;
McGillicuddy et al., 1998). Variability in the water column structure
was visible during the study period (Figure 2). This is best evidenced
through the temperature profiles; In the late summer and early fall
there was a stratified water column with high temperatures in the
surface (\textless75m) (Figure 2A) and slightly elevated salinity
(Figure 2B). This warm, stratified period appeared more intense during
the few months sampled in 2019. In 2021, we observed the stratified
layer slowly dissipated into the winter months. There was a consistent
oxygen minimum zone (OMZ) located at about 750m deep (Figure 2C).
Although, February 2021 saw a notable downwelling event, likely due to a
passing cyclonic eddy. During this time, warmer, oxic water was plunged
deeper into the mesopelagic. This process was reversed in the spring
months (March, April) when mixing occurred throughout the epipelagic and
the surface was cooler and well mixed. Primary production was highest
during the spring mixing period, evidenced both by in-situ fluorescence
(Figure 2D) and productivity incubation experiments (Figure 3A).
Originating near the surface, the productivity peak moved deeper
throughout the spring and declined into the summer (Figure 2D). However,
there was a notable, yet smaller productivity bump in the late summer
and early fall (Figure 3A) which occurred deeper in the epipelagic
(Figure 2D). The particle concentration (184\(\mu m\) - 450\(\mu m\))
was closely coupled to chlorophyll-a patterns.

\begin{figure}

{\centering \includegraphics{images/02-environmental.pdf}

}

\caption{Environmental profiles across time-series of study period. Y
axis shows depth in meters. (A) Temperature. (B) Salinity. (C) Dissolved
Oxygen. (D) In-situ chlorophyll fluorescence. (E) Particle concenctrion
(184 - 450\(\mu m\)). (F) Bacteria Abundance. (G) Silica. (H) Nitrate}

\end{figure}
